Python as a machine learning language: experiences from the scikit-learn project

### 2013-06-15 paper.tex
parallelism

Our implementation guidelines emphasize on writing short and readable code,
easily maintainable by external contributors, while being as efficient as
possible.  Whenever practicable, algorithms implemented in \sklearn are thus
directly written in pure Python. For intensive numerical computations, the
library mostly relies on NumPy vector operations, which allows for the code to
remain concise and readable while being efficient. For all algorithms that
cannot be easily and efficiently expressed as NumPy operations, we rely on
Cython \citep{behnel2011cython}  for writing the critical parts and achieving
decent performance and scalability. Cython is a compiled programming language
that extends Python with static typing. It produces efficient C extension
modules that are directly importable from the Python runtime system. Examples of
algorithms written in Cython include stochastic gradient descent for linear
models, some graph-based clustering algorithms or decision trees. To facilitate
the adoption of the library, we also limit to a bare minimum the set of external
dependencies. As such, NumPy and SciPy -- the two most common dependencies in the
Python scientific ecosystem -- are actually the only two required dependencies
for the library to fully work. Some modules may rely on other external
libraries, but then we tend to prefer to directly ship them within the codebase
rather than adding new dependencies. In particular, the codebase ships modified
versions of the \textsf{LIBSVM} and \textsf{LIBLINEAR} libraries
\citep{chang2011libsvm,fan2008}, both written in C{}\verb!++! and wrapped using
Cython modules.

@@

At this point in time, support for multi-core programming in Python,
at least in the CPython reference implementation targeted by scikit-learn,
is suboptimal.
The interpreter supports multithreading,
but due to the ``Global Interpreter Lock'' (GIL),
no two threads in a process may at the same time execute Python code.
It follows that multithreading is not a viable solution
even for embarrassingly parallel problems,
such as fitting and evaluation multiple models in a grid search procedure.
The workaround is to use Python's multiprocessing module,
which emulates threads using multiple operating system-level processes,
but this in turn leads to high memory use and overhead
because of the copying of training and test sets
into the various processes' address spaces.

>>>>

Our implementation guidelines emphasize on writing short and readable code,
easily maintainable by external contributors, while being as efficient as
possible.  Whenever practicable, algorithms implemented in \sklearn are thus
directly written in pure Python. For intensive numerical computations though,
the library mostly relies on NumPy vector operations, which allows for the code
to remain concise and readable while being efficient. For all algorithms that
cannot be easily and efficiently expressed as NumPy operations, we rely on
Cython \citep{behnel2011cython}  for writing the critical parts and achieving
decent performance and scalability. Cython is a compiled programming language
that extends Python with static typing. It produces efficient C extension
modules that are directly importable from the Python runtime system. Examples of
algorithms written in Cython include stochastic gradient descent for linear
models, some graph-based clustering algorithms or decision trees. To facilitate
the adoption of the library, we also try to limit to a bare minimum the set of
external dependencies. As such, NumPy and SciPy -- the two most common
dependencies in the Python scientific ecosystem -- are actually the only two
required dependencies for the library to fully work. Some modules may rely on
other external libraries, but in such cases we tend to prefer to directly ship
them within the codebase rather than adding new dependencies. In particular, the
codebase includes modified versions of the \textsf{LIBSVM} and
\textsf{LIBLINEAR} libraries \citep{chang2011libsvm,fan2008}, both written in
C{}\verb!++! and wrapped using Cython modules.

@@

In many cases, machine learning algorithms can be decomposed into independent
smaller tasks that may each be solved in parallel, thereby reducing the overall
computing times. Some algorithms in \sklearn are implemented as such, and
exploit multi-core architectures.

At this point in time, support for multi-core programming in Python -- at least
in the CPython implementation targeted by scikit-learn -- is however suboptimal.
The interpreter supports multithreading, but due to the \textit{global
interpreter lock}, no two threads in a process may execute Python code at the
same time. It follows that multithreading is not a viable solution for
parallelizing parts of an algorithm, even for embarrassingly parallel problems,
such as fitting and evaluating multiple models in a grid search procedure. The
workaround that we use is to rely on the Python multiprocessing module, which
emulates threads using multiple operating system-level processes. However this
solution is not perfect as it may in turn lead to a large memory overhead since
objects (e.g., input datasets) need to be copied within each process own memory
space. In some cases, it may unfortunately even happen that the overhead becomes
so significant that fully exploiting a multi-core architecture becomes
infeasible due memory constraints. As such, a proper implementation of parallel
machine learning algorithms remain a challenging task for the project.
